---
title: Data for Diamond Princess analysis
author: Alexander E. Zarebski
format:
  html:
    toc: true
knitr:
  opts_knit:
    root.dir: "~/projects/diamond-princess-timtam"
---

# Summary

- The current `diamond.fasta` file used in the analysis has 71
  sequences.
- The previous analysis lists 70 accession numbers. When I looked on
  GenBank (via `rentrez`) I could only get 69 of the sequences.
- It looks like we can safely ignore sequence `DP0321`.
- The `timtamslamR` package makes sorting out times much easier.

# Setting up

```{r}
library(purrr)
library(rentrez)
library(ape)
library(xml2)
library(stringr)
library(timtamslamR)
```

If you don't have `timtamslamR`, you can install it with devtools:

```r
devtools::install_github("aezarebski/timtamslamR")
```

# Data

```{r}
#| echo: false
## By default, `rentrez` returns fetched data as an `XML` object from
## the XML package, rather than as an `xml_node` from the `xml2`
## package, which is much easier to work with, so I'll add a helper
## here to move between them.
fetch_as_xml <- function(...) {
  temp_file <- tempfile()
  on.exit(file.remove(temp_file))

  entrez_obj <- entrez_fetch(
    ...,
    rettype = "xml",
    parsed = TRUE
  )
  XML::saveXML(entrez_obj, file = temp_file)
  return(xml2::read_xml(temp_file))
}
```

Define a couple of file paths where we can store the data. Here we
have a copy of the sequence accession numbers taken from Appendix G of
Andréoletti, Zwaans et al (2022).

```{r}
entrez_fasta <- "data/entrez-result.fasta"
entrez_xml <- "data/entrez.xml"
diamond_fasta <- "data/diamond.fasta"

accession_numbers <- c(
  "LC570961.1","LC570962.1","LC570965.1","LC570967.1","LC570968.1",
  "LC570969.1","LC570970.1","LC570971.1","LC570972.1","LC570973.1",
  "LC570974.1","LC570975.1","LC570976.1","LC570977.1","LC570978.1",
  "LC570979.1","LC570980.1","LC570981.1","LC570982.1","LC570983.1",
  "LC570984.1","LC570985.1","LC570986.1","LC570987.1","LC570988.1",
  "LC570989.1","LC570990.1","LC570991.1","LC570992.1","LC570993.1",
  "LC570994.1","LC570995.1","LC570996.1","LC570997.1","LC570998.1",
  "LC570999.1","LC571000.1","LC571002.1","LC571003.1","LC571004.1",
  "LC571005.1","LC571006.1","LC571007.1","LC571008.1","LC571009.1",
  "LC571010.1","LC571011.1","LC571012.1","LC571013.1","LC571014.1",
  "LC571017.1","LC571019.1","LC571020.1","LC571022.1","LC571023.1",
  "LC571024.1","LC571025.1","LC571028.1","LC571030.1","LC571031.1",
  "LC571032.1","LC571034.1","LC571035.1","LC571036.1","LC571037.1",
  "LC571038.1","LC571039.1","LC571040.1","LC571041.1"
)
```

## Downloading data with R

We can then download these sequences from what should be the GenBank
database using the `rentrez` package. To avoid unnecessary calls to
the database we can check if this data already exists locally.

```{r}
entrez_obj <- fetch_as_xml(db = "nucleotide", id = accession_numbers)
```

## Creating initial FASTA

We want a FASTA file that has identifiers consisting of the accession
number, the isolate name (i.e., the "DPXXXX" name) and the collection
date. The following XPath queries and the `text_from_nodes` function
simplify the task of extracting the information.

```{r}
collection_date_xp <-
  "//GBQualifier[GBQualifier_name[text()='collection_date']]/GBQualifier_value"
primary_accession_xp <- "//GBSeq_primary-accession"
isolate_xp <-
  "//GBQualifier[GBQualifier_name[text()='isolate']]/GBQualifier_value"
sequence_xp <- "//GBSeq_sequence"

text_from_nodes <- function(nodeset, xpath) {
    return(xml_text(xml_find_all(nodeset, xpath)))
}

seq_nodes <- entrez_obj |> xml_find_all("//GBSeq")

collection_dates <-
  text_from_nodes(seq_nodes, collection_date_xp)

primary_accessions <-
  text_from_nodes(seq_nodes, primary_accession_xp)

isolate_names <-
  text_from_nodes(seq_nodes, isolate_xp)

sequences <-
  text_from_nodes(seq_nodes, sequence_xp) |>
  map(tolower) |>
  map(strsplit, "") |>
  map(unlist)
```

Use a `DNAbin` to make it easier to write this to FASTA.

```{r}
dna <- as.DNAbin(sequences)
names(dna) <- str_c(primary_accessions,
                    isolate_names,
                    collection_dates,
                    sep = "_")

write.dna(dna, file = entrez_fasta, format = "fasta")
```

## Create an alignment and mask questionable parts of the genome

First we need something to align against, so we will pull the same
sequence as used before.

```r
reference_genome_xml <-
  fetch_as_xml(db = "nucleotide", id = "MN908947")
```

## Give up, cheat and use the existing alignment

TODO It would be very nice to be able to have the subsequent steps of
the original analysis included here, but until this can be
accomplished we can just cheat and rename the existing alignment.

```{r}
old_msa <- read.FASTA("data/diamond.fasta")
old_names <- names(old_msa)
new_msa <- read.FASTA(entrez_fasta)
new_names <- names(new_msa)

renamed_msa <- list()
## Use old names for iterations to keep the same order as in the old
## MSA.
for (seq_name in names(old_msa)) {
  isolate_name <- str_extract(seq_name, "DP[0-9]{4}")
  if (any(str_detect(string = new_names, pattern = isolate_name))) {
    old_ix <- which(str_detect(string = old_names,
                               pattern = isolate_name))
    new_ix <- which(str_detect(string = new_names,
                               pattern = isolate_name))
    tmp <- old_msa[old_ix]
    names(tmp) <- new_names[new_ix]
    renamed_msa <- c(renamed_msa, list(tmp))
  }
}
renamed_msa <- do.call(what = c, args = renamed_msa)

write.FASTA(renamed_msa, file = "data/cheating.fasta")
```

```{r}
#| echo: false
## Frustratingly, this has written the genomes in upper case, so we
## need to create another version with the lower case to get the same
## results as before. Put all the sequences in lower case to help
## check diff.
tmpA <- readLines("data/cheating.fasta")
is_id <- str_detect(tmpA, "^>")
tmpB <- character(length(tmpA))
for (ix in seq_along(tmpA)) {
  if (is_id[ix]) {
    tmpB[ix] <- tmpA[ix]
  } else {
    tmpB[ix] <- tolower(tmpA[ix])
  }
}
writeLines(tmpB, "data/cheating.fasta")
```

## Use timtamslamR to put time stamps on the MSA

```{r}
seqs_calendar <- read_fasta("data/cheating.fasta")
```

The figure below shows the number of sequences obtained on each date.

```{r}
#| echo: false
plot_dates(seqs_calendar)
```

We can use method A to assign uniformly times to the sequences in a
way that avoids conflicts with the timing of rate changes or the time
series of unsequenced cases.

```{r}
seqs_timed <- rename_dates_to_times_a(seqs_calendar)
write_fasta(seqs_timed, "data/time-stamped-diamond.fasta")
```

The figure below shows the new time stamps for each of the sequences.

```{r}
#| echo: false
plot_times(seqs_timed)
```

## Time stamp the time series of unsequenced cases

The time of the last sequence is needed to time stamp the time series.
We put the time series measurements at midday on the calendar date
indicated by the CSV and then adjust for the difference in the time of
the last sequenced sample which is used as the _present_.

```{r}
present_time <- get_present(seqs_calendar, seqs_timed)
time_series_calendar <- read.csv("data/time-series-cases.csv")
time_series_timed <-
  rename_time_series(present_time, time_series_calendar)
write.csv(time_series_timed, "data/time-series-cases-timed.csv")
```

# Creating BEAST2 XML

TODO Describe how to go from the (timed) sequence and time series data
to the final XML that we give to BEAST2.

<!--  LocalWords:  GenBank XPath CSV
 -->
