---
title: Data for Diamond Princess analysis
author: Alexander E. Zarebski
format:
  html:
    toc: true
knitr:
  opts_knit:
    root.dir: "~/projects/timtam-diamond-princess"
---

# Summary

- The current `diamond.fasta` file used in the analysis has 70
  `seqs_calendar` entries, which are all seqs_calendar published in
  ref: https://doi.org/10.1073/pnas.2006824117, also used by
  Andréoletti, Zwaans et al (2022), as listed in Appendix G.
- This document describes how this data is processed for analysis and
  how times are defined in the BEAST2 XML for Timtam.
- The `timtamslamR` package makes sorting out times much easier.

# Setting up

```{r}
library(purrr)
library(rentrez)
library(ape)
library(xml2)
library(stringr)
library(timtamslamR)
```

If you don't have `timtamslamR`, you can install it using the
`devtools` package:

```r
devtools::install_github("aezarebski/timtamslamR")
```

The sequencing data is obtained from GISAID's EpiCoV database
(https://gisaid.org/), and aligned using Nextstrain
https://docs.nextstrain.org/projects/ncov/en/latest/tutorial/custom-data.html.
Using a GISAID account, the dataset is easiest retrieved in EpiCoV by
searching by virus name and location:

- name: DP0 (for Diamond Princess)
- location: Japan

This search should yield 71 `seqs_calendar` records whose `epi_isl_id`
numbers are listed below:

```{r}
gisaid_ids <- c("EPI_ISL_416565","EPI_ISL_416566","EPI_ISL_416567",
                "EPI_ISL_416568","EPI_ISL_416569","EPI_ISL_416570",
                "EPI_ISL_416571","EPI_ISL_416572","EPI_ISL_416573",
                "EPI_ISL_416574","EPI_ISL_416575","EPI_ISL_416576",
                "EPI_ISL_416577","EPI_ISL_416578","EPI_ISL_416579",
                "EPI_ISL_416580","EPI_ISL_416581","EPI_ISL_416582",
                "EPI_ISL_416583","EPI_ISL_416584","EPI_ISL_416585",
                "EPI_ISL_416586","EPI_ISL_416587","EPI_ISL_416588",
                "EPI_ISL_416589","EPI_ISL_416590","EPI_ISL_416591",
                "EPI_ISL_416592","EPI_ISL_416593","EPI_ISL_416594",
                "EPI_ISL_416595","EPI_ISL_416596","EPI_ISL_416597",
                "EPI_ISL_416598","EPI_ISL_416599","EPI_ISL_416600",
                "EPI_ISL_416601","EPI_ISL_416602","EPI_ISL_416603",
                "EPI_ISL_416604","EPI_ISL_416605","EPI_ISL_416606",
                "EPI_ISL_416607","EPI_ISL_416608","EPI_ISL_416609",
                "EPI_ISL_416610","EPI_ISL_416611","EPI_ISL_416612",
                "EPI_ISL_416613","EPI_ISL_416614","EPI_ISL_416615",
                "EPI_ISL_416616","EPI_ISL_416617","EPI_ISL_416618",
                "EPI_ISL_416619","EPI_ISL_416620","EPI_ISL_416621",
                "EPI_ISL_416622","EPI_ISL_416623","EPI_ISL_416624",
                "EPI_ISL_416625","EPI_ISL_416626","EPI_ISL_416627",
                "EPI_ISL_416628","EPI_ISL_416629","EPI_ISL_416630",
                "EPI_ISL_416631","EPI_ISL_416632","EPI_ISL_416633",
                "EPI_ISL_416634","EPI_ISL_454749")

```

For further processing, the GISAID metadata file,
`custom.metadata.tsv`, as well as the alignment file called
`aligned_custom_data.fasta` produced by the ncov pipeline, are placed
in the `~/data` directory.

## Editing the initial FASTA

We want a FASTA file that has identifiers consisting of the isolate
name (i.e., the "DPXXXX" name) and the relative sequencing time
(relative to the start of the cruise forward in time).

```{r}
seqs_calendar <- ape::read.FASTA("./data/aligned_custom_data.fasta")

metadata <- read_tsv("./data/custom.metadata.tsv")

isolate_names <- gsub("Japan","",names(seqs_calendar))
isolate_names <- gsub("/","",isolate_names)
isolate_names <- gsub("2020","",isolate_names)

names <- paste0(metadata$gisaid_epi_isl,rep("/",71),isolate_names,rep("/",71),metadata$date)
names(seqs_calendar) <- names

dna <- as.alignment(seqs_calendar)
DNAbin <- as.DNAbin.alignment(dna)
names(DNAbin) <- names

```

The figure below shows the number of `seqs_calendar` obtained on each date.

```{r}
#| echo: false
plot_dates(DNAbin)
```

We can use method A to assign uniformly times to the sequences in a
way that avoids conflicts with the timing of rate changes or the time
series of unsequenced cases, and excluding sequence DP0321. Times are
uniformly distributed within each day: there are 21 sequences on 15.02
(26th day), 23 on 16.02 (27th day) and 29 on 17.02 (28th day). We save
this file as `diamond.fasta` as input for BEAST2.

```{r}
seqs_timed <- seqs_calendar
names_times <- paste0(isolate_names[1:70],rep("/",70),c(seq(26+1/21,27-1/21,by=1/21),seq(27+1/23,28-1/23,by=1/23),seq(28+1/29,29-1/29,by=1/29) ))

names(seqs_timed) <- names_times
write_fasta(seqs_timed, "./data/diamond.fasta")
```

The figure below shows the new time stamps for each of the `seqs_calendar`.

```{r}
#| echo: false
plot_times(seqs_timed)
```

## Offset of times used in the analysis to set relative times

All times in the XML are specified relative to the time of the last
sequence:

```{r}
last_seq_time <- get_present(seqs_calendar, seqs_timed)
```

In absolute time from the first day of the cruise, this is at time
28.965517. Internally, BEAST2 sets this time as 0, and all other times
are specified backwards in time in the XML. It follows that midnight
on the last day of sequencing is: 28.965517 - 29.0 = -0.034483

## Assigning times to time series events (`disasterTimes`)

We assume the population is tested at midday on each day:

Midday on the last day with a sequence: `-0.034483 + 0.5 = 0.465517`

Midday on the following day is: `-0.034483 - 0.5 = -0.534483`

## Assigning times for prevalence estimates

We estimate prevalence at 9AM on each day weekly.

With the same logic 9AM on day X in relative time corresponds to: `-0.034483 + X.625 = X.590517`

After the last sequence: `-0.034483 - X.375 = -X.409483`

## Parameter changes can happen at midnight

As shown above, midnight on any day X before the last sequence corresponds to the following: `X.965517`
